No time for dead time - Use the Fourier amplitude differences to normalize dead time-affected periodograms

### 2017-12-11 referee_response.md

> Also in this section the expected Fourier amplitude is given as N_phot P(\nu)/4, and only *later* in the paragraph is it explicitly stated as the Leahy normalization.

The section was basically rewritten.

> (As a side note, I see figures on the Jupyter notebook with the Leahy power of white noise being plotted with values of order 100, which is not what I expect for the Leahy normalization.)

We do not fully grasp this point, and we cannot find a place in the notebook where Leahy power from _white noise_ goes to 100, in which case it might just be a misprint of the axis label. The white noise distributions we plot show precisely how the data points follow the correct chi squared distribution. Moreover, the points where the Leahy power goes to 100 in notebook 2 - if the reviewer is referring to that plot - is heavily distorted by dead time and amplified (around 300 Hz is the maximum of the deadtime positive distortion).

@@

> In general, there is too much explanation and description crammed into the figure caption, instead of the text. I assume this was in some attempt to obtain a "letter" length. Given rapid electronic publication these days, there often isn't a strong motivation for choosing a letter over a journal article, especially if that means sacrificing both clarity and depth.

We respectfully disagree with this. The text in the figures counts exactly as much as the one in the main text for the ApJL rules, so there is would be no advantage whatsoever from the point of view of obtaining a letter length. We just like the figures to work on their own, without the need to look in the main text for essential information needed to understand them. The figures were quite (probably too) complicated and needed some text to describe them properly. In any case, now most of them have changed.

> And we haven't even gotten to discussing the algorithm. The one major question I have is, what happens if the two lightcurve extractions are in fact not identical? How precisely do the two detectors have to behave identically? This is mentioned at the end of Section 5, but not adequately discussed.

In the new version, we discuss more extensively the details of this, and other caveats, in a subsection on its own. Basically, the only power spectral measurement that is heavily affected by the count rate difference in the two detectors is the single-channel periodogram. Neither the periodogram obtained by the total-intensity light curve (the sum of the two channels) or the cospectrum are very sensitive to the difference of flux in the two channel. In our simulations the parameter indicating the ratio of flux in the two channels was free to change from 0.5 to 2 (with 1 being the same flux in the two channels), and we find basically indistinguishable distributions if the fluxes in the channels are within ~30% of one another (It can be seen in the simulations in the notebooks). The simulations in figure 5 include channel ratios between 0.8 to 1.3.

### 2017-12-12 referee_response.md

Report from Referee:

> At this point, I cannot recommend for publication "No Time for Dead Time - Use of the Fourier Amplitude Differences to Normalize Dead Time-Affected Periodograms" by Bachetti and Hupenkothen. The basic reason is that the paper as written is difficult enough to follow, contains typographical errors, misstatements and mislabelings, has poor figures, to the point that I cannot easily create my own code that reproduces the results exactly as presented in this work. An algorithmic paper like this, to be truly useful, should be sufficiently self-contained and explanatory so as to be easily reproducible.

We thank the referee for the frank report, that allowed to improve the paper significantly. We do acknowledge that the paper was not as clear as it should have been. However, we also do think that a poor description should not be confused with a poor result. The extensive modifications we have done to the manuscript will hopefully demonstrate this. We would like to start this response with the description of a possible critical point: in order to accommodate a thorough discussion of the caveats of the method, we decided to take out the real data example from the paper (Cyg X-1). We realize that this choice might raise some concerns from the referee, as traditionally a paper is considered incomplete without real data. However, we find that the addition of real data does not really add anything to the main focus of the paper, which is the description of a data analysis method that works on any instrument with multiple identical detectors (it will still work practically unmodified with IXPE, that will probably have a dead time only slightly shorter than NuSTAR, and of course Astrosat etc.), and will be very useful to the community. Besides the reassuring fact of seeing a real periodogram from a real BH candidate, a single example of real data spectrum would not do a better job at showing the method, because it would still be a single example, subject to the usual doubts (is it a cherry-picked example?). The simulation environment we use in the paper mimicks NuSTAR sufficiently well, uses thousands of realizations of simulated NuSTAR power spectra - not just one - and the new additions in the discussion about the difference of flux between the two channels also includes effects of count rate differences that are higher than expected from any reasonable situation found in NuSTAR data (where bright sources are usually dominating the observation and are far from the detector edges). In any case, we are setting up a web page showing off the FAD correction working on real data that we will link in the text.

@@

> I realize that the authors have provided a link to a Jupyter notebook to reproduce the results and figures, and it was sufficient to highlight some of the misstatements made in this paper. However, I have some issues with the adequacy of this, as discussed below.
> In section 2.1, the discussion on creating the lightcurves is insufficient. Yes, it is reasonable to create lightcurves with an existing set of packages, but since these packages are in active development a version number for the packages should be given, or a git hash for the versions used.
> How do I know whether any version I run now or in the future will not be different?

We do agree that it is good practice to refer to a specific version of the codes, so we added the version numbers used.

> More importantly "stingray.simulate.Simulator" class is *not* a description of the methods used. A few sentences describing the mechanics behind each method used is necessary. Likewise, going to the Jupyter notebook and simply seeing "from stingray" and "from hendrics" is not a description. Unless I start following a potentially long chain of imports and start reading code, "stingray" and "hendrics" are black boxes. (In general, the existing Jupyter notebook desperately needs comments in the code.) This can be rectified with short descriptions of the algorithms.

We have added some description of the algorithms, in some cases by referencing Bachetti+15 where some of them were already described in detail (the simulation process is basically the same: model periodogram -> Timmer & Koening method -> light curve -> acceptance-rejection method -> clean event lists -> dead time filter -> final event lists). We hope the new version will be easier to follow.

@@

As we said in the opening, we eliminated the part about NuSTAR data analysis from the paper.

> Section 3, x and y are defined, without explicitly saying that you are specifically considering binned lightcurves of identical length representing the exact same times.
We added this information

> Also in this section the expected Fourier amplitude is given as N_phot P(\nu)/4, and only *later* in the paragraph is it explicitly stated as the Leahy normalization.

The section was basically rewritten, but we did pay attention to specifying the normalization of P the first time it is mentioned.

@@

> Section 4, Figure 3 is said when you mean Figure 2, and

Corrected, and figures are changed anyway.

> Figure 2a is said to be Fourier Amplitudes vs. Fourier Amplitude Differences, when you actually mean the standard deviation of said quantities. (The Jupyter notebook was useful for figuring this out.)

No, we made a confusing mistake in the plot and the notebook and this needs to be clarified. The only point missing in the text is that the _mean absolute values_ of the FAD and amplitudes in each frequency bin are used, and this is the reason for the apparent lack of negative values. The description with the standard deviation is somehow equivalent, with actually the same factor between the quantities (we tried both approaches in the Jupyter notebook, and we chose the plots inconsistently), but it's not what we use in the actual algorithm (see the implementation in the file fad_correction.py in the notebooks/ directory in the repository).

### 2017-12-12 paper/deadtime.tex
Put everything in submittable format
Minor edits here and there

\section{On the difference of Fourier amplitudes} \label{sec:fourierdiff}
\begin{figure*}
\gridline{\fig{imgs/rn_fourierdiff.eps}{0.33\textwidth}{(a)}
	\fig{imgs/wn_fourierdiff.eps}{0.33\textwidth}{(b)}
	\fig{imgs/fourier_vs_diff_dt.eps}{0.33\textwidth}{(c)}}

@@

Let us consider two identical and independent detectors observing the same variable source, producing independent and strictly simultaneous time series, with identical binning, $\mathbf{x} = \{x_k\}_{k=1}^N$ and $\mathbf{y} = \{y_k\}_{k=1}^N$. For a stochastic process (e.g.\ $1/\nu$-type red noise), the Fourier amplitudes will vary as a function of $N_{\mathrm{phot}}P(\nu)/4$, where $P(\nu)$ (Leahy-normalized, \citealt{Leahy+83}) is the shape of the power spectrum underlying the stochastic process, and $N_{\mathrm{phot}}$ denotes the number of photons in a light curve. If the two detectors observe the same source simultaneously, the amplitudes and phases of the stochastic process will be shared among $\mathbf{x}$ and $\mathbf{y}$, while each light curve will be affected \textit{independently} by both the photon counting noise in the detector, as well as the dead time process. 
Dead time can be considered a convolution on the signal \citep{Vikhlinin+94}. 
As such, for the convolution theorem we will have that the Fourier transform $\mathcal{F}$ of dead time-affected light curves will result in a \textit{product} of the Fourier transform of the signal $\mathcal{S}$ times the Fourier transform of the dead time filter $\mathcal{D}$:
\begin{equation}
\mathcal{F}(\nu) = \mathcal{S}(\nu)\cdot\mathcal{D}(\nu)
\end{equation}
For a large enough number of data points $N$, the complex Fourier amplitudes $S_j$ will be composed of a sum of two independent random normal variables: $S_j = S_{sj} + S_{nj}$, with $\Re (S_{sj}) \sim \Normal(0, \sigma_{sj}^2)$ and $\Re(S_{nj}) \sim \Normal(0, \sigma_n^2)$, and similarly for the imaginary parts.
$\sigma_{sj}^2 = \sigma_{s}^2(\nu) = N_\mathrm{phot}P(\nu)/4$ (where $N_{\mathrm{phot}} = \sum_{k=1}^{N}{x_k}$) is given by the power spectrum of the underlying stochastic process and is frequency-dependent. However, $\sigma_n^2 = N_\mathrm{phot}/2$ is not. 
Note that $S_{xsj} = S_{ysj}$, because the amplitudes of the stationary noise process will be the same for the Fourier transforms of $\mathbf{x}$ and $\mathbf{y}$, while the components due to white noise differ between the two time series.
Moreover, the dead time filter will only depend on the mean count rate, which is again the same in the two light curves.
Thus, the difference between the Fourier amplitudes for the two time series $\mathbf{x}$ and $\mathbf{y}$ will be:
\begin{equation}
F_{xj} - F_{yj} = (S_{xj} - S_{yj})\cdot D_j = (S_{xnj} - S_{ynj})\cdot D_j
\end{equation}
So, in practice, the real and imaginary parts of the Fourier amplitude \textit{difference} between two simultaneous lightcurves of the same signal from identical detectors will be a random variate of the form $\Normal(0, 2\sigma_{n}^2)$ for both the imaginary and real parts, \textit{multiplied} by the Fourier transform of the dead time filter.
But the squared norm of the latter \textit{is also what is distorting the periodograms}!
However, in the case of periodograms the source signal is mingled with the shape of $|\mathcal{D}|^2$, while in this case we have a \textit{constant-scatter} random variable which is modulated by the same shape. Therefore, we can try to extract the shape of $|\mathcal{D}|^2$ from the Fourier amplitude differences from the two channels, and use it to correct the shape of the periodogram.

@@

\begin{figure*}
\plotone{imgs/before_and_after.eps}
%\plotone{imgs/pds_pdf.pdf}
\caption{Periodogram and cospectrum, before and after FAD correction, for a pure white noise light curve. 
The deadtime-driven distortion of the white noise level in the periodogram, and the frequency-dependent modulation of the \rms in both spectra, disappear after applying the FAD correction. 
Spectra were calculated over 2-sec intervals and averaged, to decrease the scatter and highlight the distortion of powers.}
\label{fig:comparison}
\end{figure*}

\begin{figure*}
\plotone{imgs/pds_pdf.pdf}
\caption{PDF of non-averaged powers in the cospectrum (red) and the periodogram (black), before the FAD correction and after, shown as a histogram. 
After correction, the powers follow remarkably well the expected Laplace and $\chi^2_2$ distributions respectively, as highlighted by the overplotted probability density functions (PDF).}
\label{fig:dist}
\end{figure*}

Let us simulate two constant light curves with an incident mean count rate of 400 counts/sec and a dead time of 2.5 ms, as we would expect from two identical detectors observing the same stable X-ray source.
The Fourier amplitudes of the light curves from the two detectors are heavily distorted by dead time, with the characteristic damped oscillator-like shape \citep{Vikhlinin+94,Zhang+95}  (Figure ~\ref{fig:fourierdiff}, middle panel). 
As laid out in Section \ref{sec:fourierdiff}, the difference of Fourier amplitudes from two independent but identical detectors shows no source variability, but \textit{still shows the same distortion} due to dead time.
Therefore, using the difference between the Fourier amplitudes in two detectors, we can in principle renormalize the power spectrum so that only the source variability alters its otherwise flat shape.

As shown in Figure~\ref{fig:fourierdiff} (right panel), the single-channel Fourier amplitudes are proportional to the difference of the Fourier amplitudes in different realizations, with a constant factor $1/\sqrt{2}$.
Therefore, we expect that the periodogram will be proportional to the square of the Fourier amplitude difference, divided by 2.
Let us try to \textit{divide the power spectrum by a smoothed version of the squared Fourier differences}, and multiply by 2.
For smoothing, we used a Gaussian running window with a window width of 50 bins.
Given that the initial binning had 50 bins/Hz, this interpolation allows an aggressive smoothing over bins whose y value does not change significantly.
We call this procedure the \textbf{Fourier Amplitude Difference} (hereafter FAD) \textbf{correction}.

The correction is depicted in Figure~\ref{fig:comparison}. 
Starting from a heavily distorted distribution of the powers, applying the FAD correction ``flattens'' remarkably well the white noise level of the periodogram and the distribution of the scatter of the white noise periodogram and cospectrum. 
Also, it reinstates a correct distribution of powers, following the expected $\chi^2_2$ distribution \citep{Lewin+88} very closely. 
Analogously, the corrected cospectrum will follow the expected Laplace distribution \citep{Huppenkothen & Bachetti, sub.}.
While the original dead time-affected cospectrum had a frequency-dependent modification to the \rms level, the FAD-corrected cospectrum gets back to a frequency-independent shape, like in the dead time-free case.

@@

\begin{figure*}
    \plotone{imgs/four_qpos.eps}
    \caption{Periodograms and cospectra from a simulation with four Lorentzian features (at 50, 200, 300 and 400 Hz) with 40-Hz full with at half maximum (FWHM). 
    We plotted and fitted periodograms and cospectra before and after applying the dead time filter
    The total \rms before dead time was 40\% and the incident photon flux 400 ct/s. 
    After applying a dead time of 2.5 ms like in \nustar, the ``detected'' photon flux decreased to $\sim$200 ct/s as expected. 
    As can be seen from the best-fit curves, there is no significant difference between FAD-normalized and deadtime-free periodograms and cospectra.
    For comparison, we also plot the cospectrum without FAD (pink), showing very different amplitudes for the four Lorentzians, due to dead time.
    }
    \label{fig:pds}
\end{figure*}

We are now ready to verify the last step: is the FAD-corrected power spectrum equivalent (albeit with some loss of sensitivity due to the lower number of photons) to the dead time-free power spectrum?
To test this, we produced a number of different synthetic datasets as explained in Section~\ref{sec:wndeadtime}, containing different combinations of QPOs and broadband noise components. 
We first calculated the periodogram and cospectrum of the dead time-free data, averaged over 128-s intervals.
Then, we applied a dead time filter to the event list and applied the FAD correction, as described in Section~\ref{sec:fad}

@@

\subsection{Simulation results and discussion}\label{sec:caveat}
\begin{figure*}
    \plotone{imgs/errors.pdf}
    \caption{Relative overestimation of FAD with respect to \rms, versus \rms, as calculated from the cospectrum.  
    We encoded the frequency of the feature in the color, and the incident rate in the size of the scatter points.
    From this visualization we see two regimes: below $\sim$40\% fractional \rms, the errors are dominated by statistical errors. 
    These errors will simply decrease when we average more data, as we expect from statistical errors.
    Over $\sim$40\% fractional \rms, the errors are significantly skewed towards an overestimation of the \rms, and this is 
    clearly truer when the incident rate \textit{and} the \rms are high.}
    \label{fig:errors}
\end{figure*}

>>>>

\section{On the difference of Fourier amplitudes} \label{sec:fourierdiff}
\begin{figure*}
\gridline{\fig{rn_fourierdiff.eps}{0.33\textwidth}{(a)}
	\fig{wn_fourierdiff.eps}{0.33\textwidth}{(b)}
	\fig{fourier_vs_diff_dt.eps}{0.33\textwidth}{(c)}}

@@

Let us consider two identical and independent detectors observing the same variable source, producing independent and strictly simultaneous time series, with identical binning, $\mathbf{x} = \{x_k\}_{k=1}^N$ and $\mathbf{y} = \{y_k\}_{k=1}^N$. For a stochastic process (e.g.\ $1/\nu$-type red noise), the Fourier amplitudes will vary as a function of $N_{\mathrm{phot}}P(\nu)/4$, where $P(\nu)$ (Leahy-normalized, \citealt{Leahy+83}) is the shape of the power spectrum underlying the stochastic process, and $N_{\mathrm{phot}}$ denotes the number of photons in a light curve. If the two detectors observe the same source simultaneously, the amplitudes and phases of the stochastic process will be shared among $\mathbf{x}$ and $\mathbf{y}$, while each light curve will be affected \textit{independently} by the photon counting noise in the detector, as well as the dead time process. 
Dead time can be considered a convolution on the signal \citep{Vikhlinin+94}. 
As such, for the convolution theorem we will have that the Fourier transform $\mathcal{F}$ of dead time-affected light curves will result in a \textit{product} of the Fourier transform of the signal $\mathcal{S}$ times the Fourier transform of the dead time filter $\mathcal{D}$:
\begin{equation}
\mathcal{F}(\nu) = \mathcal{S}(\nu)\cdot\mathcal{D}(\nu)
\end{equation}
For a large enough number of data points $N$, the complex Fourier amplitudes $S_j$ will be composed of a sum of two independent random normal variables: $S_j = S_{sj} + S_{nj}$, with $\Re (S_{sj}) \sim \Normal(0, \sigma_{sj}^2)$ and $\Re(S_{nj}) \sim \Normal(0, \sigma_n^2)$, and similarly for the imaginary parts.
$\sigma_{sj}^2 = \sigma_{s}^2(\nu) = N_\mathrm{phot}P(\nu)/4$ (where $N_{\mathrm{phot}} = \sum_{k=1}^{N}{x_k}$) is given by the power spectrum of the underlying stochastic process and is frequency-dependent. However, $\sigma_n^2 = N_\mathrm{phot}/2$ is not. 
Note that $S_{xsj} = S_{ysj}$, because the amplitudes of the stationary noise process will be the same for the Fourier transforms of $\mathbf{x}$ and $\mathbf{y}$, while the components due to white noise differ between the two time series.
Moreover, the dead time filter will only depend on the count rate, which is again the same in the two light curves.
Thus, the difference between the Fourier amplitudes for the two time series $\mathbf{x}$ and $\mathbf{y}$ will be:
\begin{equation}
F_{xj} - F_{yj} = (S_{xj} - S_{yj})\cdot D_j = (S_{xnj} - S_{ynj})\cdot D_j
\end{equation}
So, in practice, the real and imaginary parts of the Fourier amplitude \textit{difference} between two simultaneous lightcurves of the same signal from identical detectors will be a random variate of the form $\Normal(0, 2\sigma_{n}^2)$ for both the imaginary and real parts, \textit{multiplied} by the Fourier transform of the dead time filter.
But the squared norm of the latter \textit{is also what is distorting the periodograms}!
However, in the case of periodograms the source power spectrum is mingled with the shape of $|\mathcal{D}|^2$, while in this case we have a \textit{constant-scatter} random variable which is modulated by the same shape. Therefore, we can try to extract the shape of $|\mathcal{D}|^2$ from the Fourier amplitude differences from the two channels, and use it to correct the shape of the periodogram.

@@

\begin{figure*}
\plotone{before_and_after.eps}
%\plotone{pds_pdf.pdf}
\caption{Periodogram and cospectrum, before and after FAD correction, for a pure white noise light curve. 
The deadtime-driven distortion of the white noise level in the periodogram, and the frequency-dependent modulation of the \rms in both spectra, disappear after applying the FAD correction. 
Spectra were calculated over 2-sec intervals and averaged, to decrease the scatter and highlight the distortion of powers.}
\label{fig:comparison}
\end{figure*}

\begin{figure*}
\plotone{pds_pdf.pdf}
\caption{PDF of non-averaged powers in the cospectrum (red) and the periodogram (black), before the FAD correction and after, shown as a histogram. 
After correction, the powers follow remarkably well the expected Laplace and $\chi^2_2$ distributions respectively, as highlighted by the overplotted probability density functions (PDF).}
\label{fig:dist}
\end{figure*}

As laid out in Section \ref{sec:fourierdiff}, the difference of Fourier amplitudes from two independent but identical detectors shows no source variability, but \textit{still shows the same distortion} due to dead time (See Figure ~\ref{fig:fourierdiff}, left panel, where this is shown with red noise).
Let us simulate two constant light curves with an incident mean count rate of 400 counts/sec and a dead time of 2.5 ms, as we would expect from the two identical detectors of \nustar observing the same stable X-ray source.
The Fourier amplitudes of the light curves from the two detectors are heavily distorted by dead time, with the characteristic damped oscillator-like shape \citep{Vikhlinin+94,Zhang+95}  (Figure ~\ref{fig:fourierdiff}, middle panel). 
Therefore, using the difference between the Fourier amplitudes in two detectors, we can in principle renormalize the periodogram so that only the source variability alters its otherwise flat shape.

As shown in Figure~\ref{fig:fourierdiff} (right panel), the single-channel Fourier amplitudes are proportional on average to the difference of the Fourier amplitudes in different realizations, with a constant factor $1/\sqrt{2}$.
Therefore, we expect that the periodogram will be proportional to the square of the Fourier amplitude difference, divided by 2.
Let us try to \textit{divide the periodogram by a smoothed version of the squared Fourier differences}, and multiply by 2.
For smoothing, we used a Gaussian running window with a window width of 50 bins.
Given that the initial binning had 50 bins/Hz, this interpolation allows an aggressive smoothing over bins whose y value does not change significantly.
We call this procedure the \textbf{Fourier Amplitude Difference} (hereafter FAD) \textbf{correction}.

The results of this correction are shown in Figure~\ref{fig:comparison}. 
Starting from a heavily distorted distribution of the powers, applying the FAD correction ``flattens'' remarkably well the white noise level of the periodogram and the distribution of the scatter of the white noise periodogram and cospectrum. 
Also, it reinstates a correct distribution of powers, following the expected $\chi^2_2$ distribution \citep{Lewin+88} very closely. 
Analogously, the corrected cospectrum will follow the expected Laplace distribution (Huppenkothen \& Bachetti, sub.).
While the original dead time-affected cospectrum had a frequency-dependent modification to the \rms level, the FAD-corrected cospectrum gets back to a frequency-independent shape, like in the dead time-free case.

@@

\begin{figure*}
    \plotone{four_qpos.eps}
    \caption{Periodograms and cospectra from a simulation with four Lorentzian features (at 50, 200, 300 and 400 Hz) with 40-Hz full with at half maximum (FWHM). 
    We plotted and fitted periodograms and cospectra before and after applying the dead time filter
    The total \rms before dead time was 40\% and the incident photon flux 400 ct/s. 
    After applying a dead time of 2.5 ms like in \nustar, the ``detected'' photon flux decreased to $\sim$200 ct/s as expected. 
    As can be seen from the best-fit curves, there is no significant difference between FAD-normalized and deadtime-free periodograms and cospectra.
    For comparison, we also plot the cospectrum without FAD (pink), showing very different amplitudes for the four Lorentzians, due to dead time.
    }
    \label{fig:pds}
\end{figure*}

We are now ready to verify the last step: is the FAD-corrected periodogram equivalent (albeit with some loss of sensitivity due to the lower number of photons) to the dead time-free periodogram?
To test this, we produced a number of different synthetic datasets as explained in Section~\ref{sec:wndeadtime}, containing different combinations of QPOs and broadband noise components. 
We first calculated the periodogram and cospectrum of the dead time-free data, averaged over 128-s intervals.
Then, we applied a dead time filter to the event list and applied the FAD correction, as described in Section~\ref{sec:fad}

@@

\subsection{Simulation results and discussion}\label{sec:caveat}
\begin{figure*}
    \plotone{errors.pdf}
    \caption{Relative overestimation of FAD with respect to \rms, versus \rms, as calculated from the cospectrum.  
    We encoded the frequency of the feature in the color, and the incident rate in the size of the scatter points.
    From this visualization we see two regimes: below $\sim$40\% fractional \rms, the errors are dominated by statistical errors. 
    These errors will simply decrease when we average more data, as we expect from statistical errors.
    Over $\sim$40\% fractional \rms, the errors are significantly skewed towards an overestimation of the \rms, and this is 
    clearly truer when the incident rate \textit{and} the \rms are high, and the frequency relatively low.}
    \label{fig:errors}
\end{figure*}
